---
title: "DNA methylation and enrichment analysis"
author: "Daven"
output:
  html_document: default
  pdf_document: default
---

```{r setup}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("bsseq")   
BiocManager::install("DMRcate")

BiocManager::install('IlluminaHumanMethylationEPICanno.ilm10b4.hg19')

suppressPackageStartupMessages({
  library(epiwraps)
  library(ggplot2)
  library(AnnotationHub)
  library(ensembldb)
  library(edgeR)
  library(rtracklayer)
  library(BiocParallel)
  library(bsseq)   # Class and tools to analyze bisulfite sequencing data at CpG-level
  library(DMRcate) # Identification of differentially-methylated regions
})
```

# Part 1

## Load the data

Download:
```{r, download, eval=FALSE}
options(timeout = 6000)
download.file("https://ethz-ins.org/content/w11_practical.zip", "w11_practical.zip")
unzip("w11_practical.zip")
```

```{r}
DNAme <- readRDS("comparison/DNAme.rds")
tracks <- list(ATAC="comparison/ATAC.bw",
               H3K27ac="comparison/H3K27ac.bw",
               H3K4me3="comparison/H3K4me3.bw",
               DNAme=DNAme)
```

### Load the gene annotation data
```{r, annotation}
ah <- AnnotationHub()
ensdb <- ah[["AH89211"]] # GRCm38
```

Obtaining the promoter coordinates of chromosome 1:

```{r, load promoters}
chr1 <-  GRanges(seqnames=Rle(c("1")), 
                 ranges = IRanges(1, end=195471971))

# We define promoters as the regions +/- 200 of the TSS
promoterRegions <- promoters(ensdb, upstream=200, downstream=200,
                             filter=GRangesFilter(chr1), columns=c("gene_name"))
head(promoterRegions)
```

# Differential Methylation Testing

## Bsseq object

We will use a BSseq object that contains data from embryonic stem cells (ESC) and astrocytes.

```{r}
bs <- readRDS("ESC_astrocytes_BSseq.rds")
bs

# remove CpGs that have 0 coverage in some samples:
bs <- bs[rowSums(assay(bs,"Cov")>0)==ncol(bs),]

design <- model.matrix(~celltype, data=pData(bs)) 
design
# adapt the design for methylation data:
methdesign <- modelMatrixMeth(design)

bs <- suppressWarnings(
  BSmooth(bs, BPPARAM = MulticoreParam(4, progress = TRUE))
)
```

## Extracting average methylation in regions of interest

```{r}
# we get the chr1 promoters for this genome:
genes <- readRDS("hg19_chr1_genes.rds")
proms <- promoters(genes)
```

## Statistical testing

### Testing each CpG individually

```{r}
# CpG-level tests:
seqAnnot <- sequencing.annotate(bs, methdesign, all.cov=TRUE, coef="celltypeastrocytes")

# look at the top most significant CpGs:
res <- seqAnnot@ranges
res <- res[order(res$ind.fdr),]
head(res)
```

### Differentially-methylated regions (DMRs)

Identifying broader regions of coherent changes:

```{r}
dmrcateRes <- dmrcate(seqAnnot, C=2, min.cpgs=5)
dmrRanges <- extractRanges(dmrcateRes, genome="hg19")
saveRDS(dmrRanges, "dmr.rds")

ggplot(as.data.frame(dmrRanges), aes(meandiff, -log10(HMFDR), colour=width, size=no.cpgs)) + geom_point()

dmrRanges <- dmrRanges[order(abs(dmrRanges$meandiff), decreasing=TRUE)]
```

```{r}
dmr_chr1 <- dmrRanges[seqnames(dmrRanges) %in% c("chr1", "1")]
dmr_chr1 <- dmr_chr1[order(abs(dmr_chr1$meandiff), decreasing = TRUE)]
top5_dmr_chr1 <- head(dmr_chr1, 5)
```

```{r}
genes <- readRDS("hg19_chr1_genes.rds")
seqlevelsStyle(genes) <- seqlevelsStyle(top5_dmr_chr1)

genes_top5 <- subsetByOverlaps(genes, top5_dmr_chr1)

met_gene_body <- bsseq::getMeth(
  bs,
  regions = genes_top5,
  what = "perRegion"
)

gene_labels <- if ("symbol" %in% names(mcols(genes_top5))) {
  genes_top5$symbol
}

rownames(met_gene_body) <- make.unique(gene_labels)
```

```{r}
pheatmap(
  met_gene_body,
  name = "methylation",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  main = "Methylation of genes overlapping top 5 chr1 DMRs"
)
```

# Part 2

```{r}
da_url <- "https://www.ethz-ins.org/content/ATAC_AstroVsESC_subset.rds"
da_ranges <- readRDS(url(da_url))
da_ranges
```


```{r}
dmr_chr1 <- dmrRanges[seqnames(dmrRanges) %in% c("chr1", "1")]
dmr_chr1 <- dmr_chr1[order(abs(dmr_chr1$meandiff), decreasing = TRUE)]

seqlevelsStyle(da_ranges) <- seqlevelsStyle(dmr_chr1)

hits <- findOverlaps(dmr_chr1, da_ranges)

atac_change_col <- "logFC"

df_agree <- data.frame(
  DMR_index = queryHits(hits),
  DNAme_change = dmr_chr1$meandiff[queryHits(hits)],
  ATAC_change = mcols(da_ranges)[[atac_change_col]][subjectHits(hits)]
)

df_agree <- df_agree[complete.cases(df_agree), ]

df_agree$relationship <- ifelse(
  sign(df_agree$DNAme_change) == -sign(df_agree$ATAC_change),
  "opposite direction",
  "same direction"
)
```

```{r}
ggplot(df_agree, aes(DNAme_change, ATAC_change)) +
  geom_point(aes(colour = relationship), alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, colour = "black") +
  theme_bw() +
  labs(
    x = "DMR methylation difference",
    y = "Differential accessibility change",
    colour = "Relationship",
    title = "Agreement between methylation and accessibility changes"
  )

cor.test(
  df_agree$DNAme_change,
  df_agree$ATAC_change,
  method = "spearman"
)

table(df_agree$relationship)
prop.table(table(df_agree$relationship))
```

This overlaps DMRs with differential accessibility regions and compares methylation change with ATAC change at the same loci. Opposite-direction changes suggest expected regulatory agreement, where increased methylation corresponds to reduced accessibility, while same-direction changes suggest disagreement or more complex regulation.