---
title: "ATAC-seq assignment"
author: "Daven"
output: html_document
---

# Part 1
```{r}
suppressPackageStartupMessages({
  library(epiwraps)
  library(AnnotationHub)
  library(MotifDb)
  library(memes)
  library(universalmotif)
  library(ensembldb)
  library(ggplot2)
})

ah <- AnnotationHub(localHub=FALSE)
ensdb <- ah[["AH89211"]] # mouse ensembldb object
```


## Download the data

```{r, eval=FALSE}
download.file("https://ethz-ins.org/content/w7/atac.chr19.bam", "atac.chr19.bam", mode="wb")
download.file("https://ethz-ins.org/content/w7/atac.chr19.bam.bai", "atac.chr19.bam.bai", mode="wb")
```
## Preparing tracks

```{r}
bam <- "atac.chr19.bam"


# create a track using only nucleosome-free fragments, the number of cuts/insertion sites at each position
bam2bw(bam, output_bw = "NF_cuts.bw", paired=TRUE, binWidth=1L, type="ends", minFragLength=30, 
       maxFragLength=120, trim=4L, extend=3L, forceSeqlevelsStyle = "Ensembl")

# create a track using only the (10bp) centers of mono-nucleosome fragments
bam2bw(bam, output_bw = "mono_centers.bw", paired=TRUE, binWidth=5L, minFragLength=140, trim=4L,
       maxFragLength=220, type="center", extend=10L, forceSeqlevelsStyle = "Ensembl")
```

## Obtaining the sites with a CTCF motif

```{r}
# get CTCF motif
motif <- MotifDb::query(MotifDb, c("CTCF","Mus"))[[1]]
motif2 <- convert_motifs(motif, class="TFBSTools-PFMatrix")
genome <- ah[["AH68356"]]
# get the sequence for chr19:
chr19 <- import(genome)["19"]

# find motif matches across chr19
moi <- motifmatchr::matchMotifs(motif2, chr19, out="positions", p.cutoff=1e-5)[[1]]
# convert to GRanges
moi <- as(setNames(moi,names(chr19)), "GRanges")
```

## Plotting signal around the motif occurences

```{r, fig.width=8, fig.height=4}
tracks <- c(
  NF_cuts = "NF_cuts.bw",
  mono_centers = "mono_centers.bw"
)
names(tracks) <- gsub("\\.bw","",basename(tracks))

# extract signals around the motif occurences
# we zoom in to 300bp around the motif centers, in windows of 5bp
sm <- signal2Matrix(tracks, moi, w=5, extend=300)

# plot the signals:
plotEnrichedHeatmaps(sm, trim=0.95, multiScale = TRUE, axis_name = "0")
```


## Summarizing the signals

```{r}
d <- meltSignals(sm[, c("NF_cuts", "mono_centers")])

ggplot(d, aes(x = position, y = mean, colour = sample)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey40") +
  labs(
    x = "Distance from CTCF motif center (bp)",
    y = "Mean signal",
    colour = NULL,
    title = "ATAC-seq signal around high-confidence CTCF motifs on chr19",
    subtitle = "Nucleosome-free cuts and mono-nucleosome fragment centers"
  ) +
  theme_classic(base_size = 12)
```


# Part 2
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("BSgenome.Mmusculus.UCSC.mm10")

BiocManager::install("plger/betterChromVAR")

suppressPackageStartupMessages({
  library(motifmatchr)
  library(MotifDb)
  library(universalmotif)
  library(ggplot2)
  library(sechm)
  library(BSgenome.Mmusculus.UCSC.mm10)
  library(limma)
  library(betterChromVAR)
})

ah <- AnnotationHub(localHub=TRUE)
ensdb <- ah[["AH89211"]] # mouse ensembldb object
```

## Download the data

```{r, eval=FALSE}
download.file("https://ethz-ins.org/content/mouse_mm38_hippocampus.peakCounts.SE.rds", "peakCounts.rds", mode="wb")
```

```{r}
se <- readRDS("peakCounts.rds")
```

## Annotate regions

```{r}
se <- GenomeInfoDb::keepStandardChromosomes(se, pruning.mode = "coarse")
seqlevelsStyle(se) <- "UCSC"
se <- addGCBias(se, genome=BSgenome.Mmusculus.UCSC.mm10)
```

### Motif matches

```{r}
motifs <- query(MotifDb, "HOCOMOCOv13")
```

convert the motifs to something that `motifmatchr` accepts

```{r, message=FALSE}
motifs <- do.call(TFBSTools::PWMatrixList, convert_motifs(motifs, class="TFBSTools-PWMatrix"))
names(motifs) <- make.unique(sapply(motifs, \(x) gsub("\\..+","",x@name)))
```

scan regions for matches of the motifs

```{r}
matches <- motifmatchr::matchMotifs(motifs, se, genome=BSgenome.Mmusculus.UCSC.mm10)
matches
```

a SummarizedExperiment, where columns are the motifs and the rows are the regions, and the assay is a logical sparse matrix:

```{r}
assay(matches)[1:10,1:20]
```

## Get motifs deviations

```{r}
dev <- betterChromVAR(se, matches)
dev
```

The output is a SummarizedExperiment, this time with samples as columns and motifs as rows.
The values represent the relative accessibility of each motif in each sample.

## Differential analysis

We can use these motif deviations to perform differential analysis:

```{r}
dev$sex <- ifelse(grepl("female", colnames(dev)), "Female",
           ifelse(grepl("male", colnames(dev)), "Male", NA))
design <- model.matrix(~sex, data=as.data.frame(colData(dev)))
fit <- eBayes(lmFit(assay(dev, "z"), design))
res <- as.data.frame(topTable(fit, number = Inf))
res$motif <- row.names(res)
ggplot(res, aes(logFC, -log10(adj.P.Val), label=motif)) + geom_text()
```
```{r}
dev$sex
```


## Heatmap
```{r}
top_n <- 15
top_motifs <- rownames(res)[seq_len(min(top_n, nrow(res)))]

ord <- order(dev$sex)

sechm(
  dev[, ord],
  features = top_motifs,
  assayName = "z",
  gaps_at = "sex",
  cluster_cols = FALSE
)
```


